Grammar-Based Concept Alignment for Domain-Specific Machine Translation

### 2021-05-16 paper/paper.tex
reviewed up to divergences

% importance of lexica in DSMT
However, lexical exactness is, in this context, just as important as grammaticality. 
An important part of the design of a Controlled Natural Language (CNL) is the creation of a high-quality translation lexicon, preserving both semantics and grammatical correctness.

% how it is done right now and issues
A translation lexicon is often built manually, which is a time consuming task requiring significant linguistic knowledge. 
When the task is based on a corpus of parallel example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques \cite{brown-etal-1993-mathematics, och-ney-2000-improved, dyer-etal-2013-simple}. 
None of them is, however, suitable for the common case in which only a small amount of example data is available --- typically, with just one occurrence of each relevant lexical item.

@@

% additional advantage 2: morphology
Concepts are to be conceived as \textit{lemmas} equipped with morphological variations, instead of fixed word forms or phrases.
This  allows us to generate translation lexica complete with grammatical category and inflection so that correct target language forms can be selected in each syntactic context.

@@

% contextualizing our proposal
As stated in the Introduction, most existing CE solutions are based on statistical approaches and are, as a consequence, unsuitable for small datasets. 
Grammar-based approaches, making use of parallel treebanks and collectively referred to as \textit{tree-to-tree alignment methods}, have also been proposed \cite{tiedemann2011bitext}, but have historically suffered from the inconsistencies between the formalisms used to define the grammars of different languages and from the lack of robustness of parsers.
This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Grammatical Framework (GF) \cite{ranta-2011} and Universal Dependencies (UD) \cite{ws-2019-universal}.

@@

% role of UD and GF
While both formalisms independently solve the issues related to having to work with grammars that are inconsistent with each other, UD is especially appealing since, dependency trees being an easier target, several robust parsers, such as \cite{straka-etal-2016-udpipe} and \cite{chen-manning-2014-fast} are available.
Alone, UD trees are sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a morphologically-aware translation lexicon for a generative grammar. 
This is where GF comes into play: after correspondences are inferred from a parallel text, our system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it immediate to carry out small-scale translation experiments using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL), which covers the morphology and basic syntax of over 30 languages.
This is enabled by \texttt{gf-ud}, a conversion tool described in \cite{kolachina-ranta-2016-abstract} and \cite{ranta-kolachina-2017-universal}.
Concretely, then, the system we propose consists of a UD parser, an alignment module based on UD tree comparison and a program, based on \texttt{gf-ud}, that converts them into the rules of a GF translation lexicon.

@@

% algorithm description
As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e. if they match the alignment criteria. 
In the case of the example criterion discussed above, this means that their roots are labelled the same.
If this is the case, they are added to a collection of alignments, which are represented as pairs of UD (sub)trees associated with some metadata, such as the id of the sentence they were extracted from.
Such a collection is what the function will return after aligning all the dependency subtrees.
The same procedure is applied recursively to all pairs of immediate subtrees of each sentence, until the leaves are reached or alignment is no longer possible due to lack of matching criteria.
Subtrees are sorted based on their dependency label to give higher priority to pairs whose heads have the same label (cf. \textsc{sort} in the pseudocode).

@@

% iterations
When working on multiple sentences, the algorithm can be applied in an iterative fashion, so that knowledge gathered when a sentence pair is aligned can be used when working on later sentences and also to keep track of the number of occurrences of each alignment throughout the entire text.

@@

\paragraph{Matching UD labels}
The most obvious, but also most effective idea is to determine alignability based on comparing the dependency labels of the candidate UD tree pair. 
In particular, according to this idea, two subtrees in \textit{matching context}, i.e. attached to aligned heads, constitute an alignment if their roots share the same dependency label, meaning that they are in the same syntactic relation with their heads.
Note that, since the root of UD trees is always attached to a a fake node with an arc labelled \texttt{root}, this criterion also implies that full sentences are always considered to align with each other.
This is desirable since we assume that the parallel texts that are fed to our program are sentence-aligned.

\paragraph{Part-Of-Speech equivalence}
% actual POS-equiv par
As noted above, the CoNNL-U notation provides information on the grammatical categories of each word, represented as Universal POS tags \cite{petrov-etal-2012-universal}. 
Intuitively, if the nodes of two trees in matching contexts have the same POS tags, the two trees are more likely to correspond to each other than if not. 
This is especially true if we focus, for instance, solely on the open class words (defined as in the UD documentation \footnote{\url{universaldependencies.org/u/pos/all.html}}), thus ignoring prepositions, determiners, auxiliary verbs and all function words, which tend to behave differently across different languages.
A useful relation to define between dependency trees is, then, that of \textit{POS-equivalence}: two dependency trees $t$, $u$ are POS-equivalent if $M_1 = M_2 \neq \emptyset$, where $M_i$ is defined as the multiset of POS tags of all the open class word nodes of $t_i$.

@@

% examples
For instance, \textit{categorial divergences} occur when the POS tag of a word in the source language changes in its translation.
A ubiquitous example of is that of adverbial modifiers (with the UD label \texttt{advmod}) are translated prepositional phrases (with the UD label \texttt{obl}, for \textit{oblique}), such as in the English-Italian pair $\langle$\textit{She studies \textbf{consistently}, Lei studia \textbf{con costanza}}$\rangle$.
\textit{Structural divergences}, where a direct object in one language is rendered as an oblique in the other, as in the English-Swedish pair $\langle$\textit{I told \textbf{him}, Jag berättade \textbf{för honom}.}$\rangle$), are also frequently encountered.

>>>>

% lexica in DSMT
However, lexical exactness is, in this context, just as important as grammaticality. 
An important part of the design of a Controlled Natural Language (CNL) is the creation of a high-quality translation lexicon, preserving both semantics and grammatical correctness.
A translation lexicon is often built manually, which is a time consuming task requiring significant linguistic knowledge. 
When the task is based on a corpus of parallel example sentences, part of this process can be automated by means of statistical word and phrase alignment techniques \cite{brown-etal-1993-mathematics, och-ney-2000-improved, dyer-etal-2013-simple}. 
None of them is, however, suitable for the common case in which only a small amount of example data is available --- typically, with just one occurrence of each relevant lexical item.

@@

% additional advantage 2: morphology
Conceiving conepts as \textit{lemmas} equipped with morphological variations rather than fixed word forms or phrases allows us to generate translation lexica complete with grammatical category and inflection, so that correct target language forms can be selected in each syntactic context.

@@

% contextualizing our proposal
As stated in the Introduction, most existing alignment solutions are based on statistical approaches and are, as a consequence, unsuitable for small datasets. 
Grammar-based approaches, making use of parallel treebanks and collectively referred to as \textit{tree-to-tree alignment methods}, have also been proposed \cite{tiedemann2011bitext}, but have historically suffered from the inconsistencies between the formalisms used to define the grammars of different languages and from the lack of robustness of parsers.
This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Grammatical Framework (GF) \cite{ranta-2011} and Universal Dependencies (UD) \cite{ws-2019-universal}.

@@

% role of UD and GF
While both formalisms independently solve the issues related to having to work with grammars that are inconsistent with each other, UD is especially appealing since, being dependency trees an easier target, several robust parsers, such as \cite{straka-etal-2016-udpipe} and \cite{chen-manning-2014-fast} are available.
Alone, UD trees are sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a morphologically-aware translation lexicon for a generative grammar. 
This is where GF comes into play: after correspondences are inferred from a parallel text, our system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it immediate to carry out small-scale translation experiments using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL), which covers the morphology and basic syntax of over 30 languages.
This is enabled by \texttt{gf-ud}, a conversion tool described in \cite{kolachina-ranta-2016-abstract} and \cite{ranta-kolachina-2017-universal}.
Concretely, then, the system we propose consists of a UD parser, an alignment module based on UD tree comparison and a program, based on \texttt{gf-ud}, that converts them into the rules of a GF translation lexicon.

@@

% algorithm description
As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e. if they match one or more alignment criteria. 
In the case of the example criterion discussed above, this means that their roots are labelled the same.
If this is the case, they are added to a collection of alignments, which are represented as pairs of UD (sub)trees associated with some metadata, such as the id of the sentence they were extracted from.
Such a collection is what the function will return after aligning all the dependency subtrees.
The same procedure is applied recursively to all pairs of immediate subtrees of each sentence, until the leaves are reached or alignment is no longer possible due to lack of matching criteria.
Subtrees are sorted based on their dependency label to give higher priority to pairs whose heads have the same label (cf. \textsc{sort} in the pseudocode).

@@

% iterations
When working on multiple sentences, the algorithm can be applied in an iterative fashion, so that knowledge gathered when a sentence pair is aligned can be used when working on later sentences and to keep track of the number of occurrences of each alignment throughout the entire text.

@@

\paragraph{Matching UD labels}
The most obvious, but also most effective idea is to determine alignability based on comparing the dependency labels of the candidate UD tree pair. 
In particular, according to this idea, two subtrees in \textit{matching context}, i.e. attached to aligned heads, constitute an alignment if their roots share the same dependency label, meaning that they are in the same syntactic relation with their heads.
Note that, since the root of a UD tree is always attached to a a fake node with an arc labelled \texttt{root}, this criterion also implies that full sentences are always considered to align with each other.
This is desirable since we assume that the parallel texts that are fed to our program are sentence-aligned.

\paragraph{Part-Of-Speech equivalence}
% actual POS-equiv par
As noted above, the CoNNL-U notation provides information on the grammatical categories of each word, represented as Universal POS tags \cite{petrov-etal-2012-universal}. 
Intuitively, if the nodes of two trees in matching contexts have the same POS tags, the two trees are more likely to correspond to each other than if not. 
This is especially true if we focus, for instance, solely on the open class words (defined as in the UD documentation\footnote{\url{universaldependencies.org/u/pos/all.html}}), thus ignoring function words such as prepositions, determiners and auxiliary verbs, which tend to behave differently across different languages.
A useful relation to define between dependency trees is, then, that of \textit{POS-equivalence}: two dependency trees $t$, $u$ are POS-equivalent if $M_1 = M_2 \neq \emptyset$, where $M_i$ is defined as the multiset of POS tags of all the open class word nodes of $t_i$.

@@

% examples
For instance, \textit{categorial divergences} occur when the POS tag of a word in the source language changes in its translation.
A ubiquitous example of is that of adverbial modifiers (with the UD label \texttt{advmod}) translated as prepositional phrases (with the UD label \texttt{obl}, for \textit{oblique}), such as in the English-Italian pair $\langle$\textit{She studies \textbf{consistently}, Lei studia \textbf{con costanza}}$\rangle$.
\textit{Structural divergences}, where a direct object in one language is rendered as an oblique in the other, as in the English-Swedish pair $\langle$\textit{I told \textbf{him}, Jag berättade \textbf{för honom}.}$\rangle$), are also frequently encountered.
